\bf Comparing the Power of Plot Designs to Reveal Correlation

### 2016-01-08 paper/paper.Rnw
equations and little things

\begin{document}
\SweaveOpts{concordance=TRUE}

@@

$$Power_{ijklt} = \psi\left\{ \beta_{0} + \beta_{1} x_{1,i}  + \beta_{2} x_{2,j}  + \beta_{3} x_{3,k} + \beta_{4} x_{4,t} + \beta_{5} x_{1,i} \* x_{2,j} + \beta_{6} x_{5,l}(\beta_{7} + \beta_{8} x_{1,i}  + \beta_{9} x_{2,j}  + \beta_{10} x_{3,k} + \beta_{11} x_{4,t} + \beta_{12} x_{1,i} \* x_{2,j}) \gamma z_{s} + \epsilon_{ijkt} \right\}$$

@@

Table \ref{tab1} summarises the model estimates. It can be confusing the interpret the model estimates, and at first glance it appears the negative plot coefficient implies that overliad lines are substantially more powerful than the scatter plot, but this is misleading and interpreting the full model involves combining the estimates for the different interactions with abs($\rho$) and sign($\rho$). abs($\rho$) models the power difference between low and high correlations, and abs($\rho)*sign($\rho$) models an increase in power appearing in positive but not negative correlation. 
Taking these into account, the scatter plot shows superior performance and higher values of abs($\rho$), It needs to be restated that we only tested correlations at  0.3, 0.5, 0.7 and 0.9, so the results below an absolute value of 0.3 are unreliable and may be artifacts of regression smoothing methods.

@@

\begin{figure}[htbp]
\centering
\label{random2}
<<random2, dependson='random', fig.height=6, fig.width=6, out.width='0.7\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
turk.correct <- summarise(group_by(full.sub, pic_id), correct=sum(correct), attempts=sum(weights*attempts))
turk.correct$percent <- turk.correct$correct/turk.correct$attempts
ability <- merge(full.sub[,c("subj_id", "pic_id", "correct")], turk.correct[c("pic_id", "percent")], by="pic_id")
ability <- summarise(group_by(ability, subj_id), actual=sum(correct), estimated=sum(percent))
ability$diff <- ability$actual - ability$estimated
ability.re <- merge(ability, randomeffects, by="subj_id")
ggplot(data=ability.re, aes(x=diff, y=re)) + geom_point() + labs{x="Visual Ability", y="Random Effects"}
@
\caption{Visual ability plotted against random effects. The diagonal pattern suggests a strong positive relationship, indicating that the random effects in the model are a suitable measure of visual ability.}
\end{figure}

@@

<<timefit, dependson='powerfit', echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
#timelmer <-  lmer(log_time ~ smoothed + r + n + rsgn*r + plot + plot*smoothed 
                  + plot*r + plot*n + plot*rsgn*r + plot*correct + (1|subj_id), data=full.sub)
#time.fit <- summary(timelmer)$coefficients
#pv <- function(x){
#  p <- 2*(1-pt(x, 1835))
#}
#p.value <- sapply(time.fit[,3], pv)
#time.fit <- cbind(time.fit, p.value)
time.fit <- read.csv("sas/time.csv")
time.fit <- time.fit[c(1, 5, 6, 2, 3, 4, 7, 8, 12, 13, 9, 10, 11, 14),] #separate into main effects then interaction effects in same order
rownames(time.fit) <- c("Intercept", "Smoothed", "t", "r", "sign(r)", "r*sign(r)", "detected", "Plot", "Plot*Smoothed", "Plot*t", "Plot*abs(r)", "Plot*sign(r)", "Plot*abs(r)*sign(r)", "Plot*detected")
time.fit$Effect <- NULL
colnames(time.fit) <- c("Estimate", "standard error", "df", "t-value", "p-value")
kable(time.fit[1:14,], format="latex", digits=2)
#kable(time.fit[8:14,], format="latex", caption="Interaction Effects", digits=2)
@
\end{table}

@@

\begin{table}
\lab{tab3}
\centering
<<answertables, dependson='setup', echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
turk.tbl <- select(Turk, test_param, smoothed, r, n, correct, attempt)
turk.tbl$result <- paste(round(turk.tbl$correct, 0), turk.tbl$attempt, sep="/")
turk.tbl <- turk.tbl[order(turk.tbl$r, turk.tbl$n),]
scatter.uns <- data.frame(c(12, "", "", 24, "", "", 48, "", "", 96, "", ""), matrix(subset(turk.tbl, smoothed==0 & test_param=="scatter")$result, ncol=8))
scatter.sm <- data.frame(c(12, "", "", 24, "", "", 48, "", "", 96, "", ""), matrix(subset(turk.tbl, smoothed==1 & test_param=="scatter")$result, ncol=8))
line.uns <- data.frame(c(12, "", "", 24, "", "", 48, "", "", 96, "", ""), matrix(subset(turk.tbl, smoothed==0 & test_param=="line")$result, ncol=8))
line.sm <- data.frame(c(12, "", "", 24, "", "", 48, "", "", 96, "", ""), matrix(subset(turk.tbl, smoothed==1 & test_param=="line")$result, ncol=8))
colnames(scatter.uns) <- c("", "-0.9", "-0.7", "-0.5", "-0.3", "0.3", "0.5", "0.7", "0.9")
colnames(scatter.sm) <- c("","-0.9", "-0.7", "-0.5", "-0.3", "0.3", "0.5", "0.7", "0.9")
colnames(line.uns) <- c("","-0.9", "-0.7", "-0.5", "-0.3", "0.3", "0.5", "0.7", "0.9")
colnames(line.sm) <- c("","-0.9", "-0.7", "-0.5", "-0.3", "0.3", "0.5", "0.7", "0.9")
kable(scatter.uns, format="latex", caption="Unsmoothed scatter plot results")
kable(scatter.sm, format="latex", caption="Smoothed scatter plot results")
kable(line.uns, format="latex", caption="Unsmoothed overlaid line results")
kable(line.sm, format="latex", caption="Smoothed overlaid line results")
@
\end{table}

@@

\begin{figure}[htbp]
\lab{map}
\centering
<<map, , dependson='setup', echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
location <- read.csv("location.csv")
map.dat <- map_data("world")
ggplot() + geom_polygon(aes(long,lat, group=group), fill="grey65", data=map.dat) + theme_bw() +
  geom_point(data=location, aes(x=lon, y=lat, colour= "blue")) +scale_colour_manual(values=c("#FF5500", "#00FFFF")) + theme(axis.text=element_blank(), axis.title=element_blank(), legend.position="none")
@
\end{figure}

>>>>

\begin{document}
%\SweaveOpts{concordance=TRUE}

@@

\begin{eqnarray*} 
Power_{ijklt} &=& \psi \{ \beta_{0} + \beta_{1} x_{1,i}  + \beta_{2} x_{2,j}  + \beta_{3} x_{3,k} + \beta_{4} x_{4,t} + \\
 & & \beta_{5} x_{1,i} \* x_{2,j} + \beta_{6} x_{5,l}(\beta_{7} + \beta_{8} x_{1,i}  + \beta_{9} x_{2,j}  + \beta_{10} x_{3,k} + \\
 & & \beta_{11} x_{4,t} + \beta_{12} x_{1,i} \* x_{2,j}) \gamma z_{s} + \epsilon_{ijkt} \}
\end{eqnarray*}

@@

Table \ref{tab1} summarises the model estimates. It can be confusing the interpret the model estimates, and at first glance it appears the negative plot coefficient implies that overliad lines are substantially more powerful than the scatter plot, but this is misleading and interpreting the full model involves combining the estimates for the different interactions with abs($\rho$) and sign($\rho$). abs($\rho$) models the power difference between low and high correlations, and $\mbox{abs}(\rho)*\mbox{sign}(\rho$) models an increase in power appearing in positive but not negative correlation. 
Taking these into account, the scatter plot shows superior performance and higher values of abs($\rho$), It needs to be restated that we only tested correlations at  0.3, 0.5, 0.7 and 0.9, so the results below an absolute value of 0.3 are unreliable and may be artifacts of regression smoothing methods.

@@

\begin{figure}[htbp]
\centering
\label{random2}
<<random2, dependson='random', fig.height=6, fig.width=6, out.width='0.7\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
turk.correct <- summarise(group_by(full.sub, pic_id), correct=sum(correct), attempts=sum(weights*attempts))
turk.correct$percent <- turk.correct$correct/turk.correct$attempts
ability <- merge(full.sub[,c("subj_id", "pic_id", "correct")], turk.correct[c("pic_id", "percent")], by="pic_id")
ability <- summarise(group_by(ability, subj_id), actual=sum(correct), estimated=sum(percent))
ability$diff <- ability$actual - ability$estimated
ability.re <- merge(ability, randomeffects, by="subj_id")
ggplot(data=ability.re, aes(x=diff, y=re)) + geom_point() + labs(x="Visual Ability", y="Random Effects")
@
\caption{Visual ability plotted against random effects. The diagonal pattern suggests a strong positive relationship, indicating that the random effects in the model are a suitable measure of visual ability.}
\end{figure}

@@

<<timefit, dependson='powerfit', echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
#timelmer <-  lmer(log_time ~ smoothed + r + n + rsgn*r + plot + plot*smoothed 
#                  + plot*r + plot*n + plot*rsgn*r + plot*correct + (1|subj_id), data=full.sub)
#time.fit <- summary(timelmer)$coefficients
#pv <- function(x){
#  p <- 2*(1-pt(x, 1835))
#}
#p.value <- sapply(time.fit[,3], pv)
#time.fit <- cbind(time.fit, p.value)
time.fit <- read.csv("sas/time.csv")
time.fit <- time.fit[c(1, 5, 6, 2, 3, 4, 7, 8, 12, 13, 9, 10, 11, 14),] #separate into main effects then interaction effects in same order
rownames(time.fit) <- c("Intercept", "Smoothed", "t", "r", "sign(r)", "r*sign(r)", "detected", "Plot", "Plot*Smoothed", "Plot*t", "Plot*abs(r)", "Plot*sign(r)", "Plot*abs(r)*sign(r)", "Plot*detected")
time.fit$Effect <- NULL
colnames(time.fit) <- c("Estimate", "standard error", "df", "t-value", "p-value")
kable(time.fit[1:14,], format="latex", digits=2)
#kable(time.fit[8:14,], format="latex", caption="Interaction Effects", digits=2)
@
\end{table}

@@

\begin{table}
\label{tab3}
\centering
<<answertables, dependson='setup', echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
turk.tbl <- select(Turk, test_param, smoothed, r, n, correct, attempt)
turk.tbl$result <- paste(round(turk.tbl$correct, 0), turk.tbl$attempt, sep="/")
turk.tbl <- turk.tbl[order(turk.tbl$r, turk.tbl$n),]
scatter.uns <- data.frame(c(12, "", "", 24, "", "", 48, "", "", 96, "", ""), matrix(subset(turk.tbl, smoothed==0 & test_param=="scatter")$result, ncol=8))
scatter.sm <- data.frame(c(12, "", "", 24, "", "", 48, "", "", 96, "", ""), matrix(subset(turk.tbl, smoothed==1 & test_param=="scatter")$result, ncol=8))
line.uns <- data.frame(c(12, "", "", 24, "", "", 48, "", "", 96, "", ""), matrix(subset(turk.tbl, smoothed==0 & test_param=="line")$result, ncol=8))
line.sm <- data.frame(c(12, "", "", 24, "", "", 48, "", "", 96, "", ""), matrix(subset(turk.tbl, smoothed==1 & test_param=="line")$result, ncol=8))
colnames(scatter.uns) <- c("", "-0.9", "-0.7", "-0.5", "-0.3", "0.3", "0.5", "0.7", "0.9")
colnames(scatter.sm) <- c("","-0.9", "-0.7", "-0.5", "-0.3", "0.3", "0.5", "0.7", "0.9")
colnames(line.uns) <- c("","-0.9", "-0.7", "-0.5", "-0.3", "0.3", "0.5", "0.7", "0.9")
colnames(line.sm) <- c("","-0.9", "-0.7", "-0.5", "-0.3", "0.3", "0.5", "0.7", "0.9")
kable(scatter.uns, format="latex", caption="Unsmoothed scatter plot results")
kable(scatter.sm, format="latex", caption="Smoothed scatter plot results")
kable(line.uns, format="latex", caption="Unsmoothed overlaid line results")
kable(line.sm, format="latex", caption="Smoothed overlaid line results")
@
\end{table}

@@

\begin{figure}[htbp]
\label{map}
\centering
<<map, , dependson='setup', echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
location <- read.csv("location.csv")
map.dat <- map_data("world")
ggplot() + geom_polygon(aes(long,lat, group=group), fill="grey65", data=map.dat) + theme_bw() +
  geom_point(data=location, aes(x=lon, y=lat, colour= "blue")) +scale_colour_manual(values=c("#FF5500", "#00FFFF")) + theme(axis.text=element_blank(), axis.title=element_blank(), legend.position="none")
@
\end{figure}
